Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-12-11 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. @odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region. Following this criteria, a minimum of 10 years of record was chosen. Any years missing more than 30 days of streamflow were removed from the possible record period. This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states. Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].

XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction. Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior. The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output. The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble. The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors). Certain predictors were subdivided according to their areal coverage of each basin (e.g. land cover was divided into 16 subdivisions).

|              | Variable                   | Source | Geoprocessing          |
|--------------|----------------------------|--------|------------------------|
| Hydroclimate | Precipitation              | PRISM  | Basin average          |
|              | Mean Temperature           | PRISM  | Basin average          |
|              | Average Evapotranspiration | MODIS  | Basin average          |
| Geospatial   | Elevation                  | DEM    | Basin average          |
|              | Area                       | DEM    | Basin average          |
|              | Slope                      | DEM    | Basin average          |
|              | Aspect                     | DEM    | Percent areal coverage |
|              | Land Cover                 | NLCD   | Percent areal coverage |
|              | Hydrologic Soil Group      | SSURGO | Percent areal coverage |
|              | Geology                    | USGS   | Percent areal coverage |
|              | Karst                      | USGS   | Percent areal coverage |

: Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}

@@

Statisical analyses were performed on annual BFI and base flow values of instrumented streamgages. We examined temporal trends using the Mann-Kendall nonparametric trend test to determine the presence of monotonic trends [@kendall1970; @mann1945nonparametric]. The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

Trends in BFI and base flow were produced using a Mann-Kendall test for monotonic trends. This test expects that the data is non-parametric and that the observations are not autocorrelated over time. Observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI. Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).

Trends with a $\rho \le 0.05$ are considered significant.

@@

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. This region also typifies a pattern of seen along most major rivers in the study area: upstream reaches indicate higher BFI values while the downstream reaches display lower BFI values. This is assumed to be due to greater groundwater-surface water interaction at the headwaters of streams (likely due to spring outlets) and the dilution of base flow while moving down stream (**CITATION**). This pattern is seen along the Gila River, Verde River, and Little Colorado Rivers in Arizona.

@@

@@

>>>>

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction. Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree. Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model. The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble. In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors). Certain features were further subdivided according to their areal coverage within each basin (e.g. land cover was divided into 16 subdivisions). This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.

|              | Variable                  | Source       | Geoprocessing          |
|------------------|--------------------|------------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting. In this approach, the data is randomly divided into $k$ folds of equal size. The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set. This process is repeated $k$ times, with each fold serving as the validation set exactly once. Each iteration trains an independent model with the same hyperparameters but using a different subset of training data. By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy. Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation. RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.

#### Feature Selection

Machine learning models are prone to overfitting, especially when provided with a large set of predictive features (CITATION). Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017]. Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.

Feature extraction involves transforming the original dataset into a lower-dimensional feature space. However, this process generates new features that often lack the physical interpretability of the original variables. In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability. In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.

To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors). A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features. Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019]. In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].

The ten most important features (\@suppfig-shap_values) were selected based on their SHAP values and used to train a subsequent model. This refined model demonstrated improved performance and reduced computational time compared to the initial model. The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.

@@

Statistical analyses were conducted on annual BFI and base flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

@@

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable trend emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, often influenced by spring outlets, and the dilution of base flow as water moves downstream (**CITATION**). This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

@@

Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset. The overall Root Mean Square Error (RMSE) is 0.129 and the overall percent bias (pbias) is -5.6. Model performance metrics across various classifications are summarized in @tbl-performance. These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications. However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @suppfig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The significance of elevation likely stems from its cumulative effects, such as lower temperatures, increased orographic precipitation, and prolonged snowpack persistence at higher altitudes. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

Land cover and land use predictors also play a significant role in BFI estimation. Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence. Similarly, hydrologic soil types show distinct trends in their impact on BFI. Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI. In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009]. The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships. For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation. Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
